Fast, Accurate, and Scalable Genomic Structural Variation Detection on the Cloud: MapReduce based {\em Cloudbreak}

### 2013-07-03 manuscript/nm.tex
kemal instroduction

Genomic structural variations such as deletions, insertions, and inversions of DNA are widely prevalent in human populations and account for the majority of the bases that differ among normal human genomes \autocite{Mills:2011p1611, Conrad:2010ja}. However, detection of structural variations with current high-throughput sequencing technology remains a difficult problem, with limited concordance between available algorithms and high false discovery rates \autocite{Mills:2011p1611}.

@@

Google's MapReduce \autocite{Dean:2008p277} framework was designed to manage the storage and efficient processing of very large scale data sets across clusters of commodity servers. Hadoop is an open source project of the Apache Foundation which provides an implementation of the MapReduce programming framework as well as a distributed file system (HDFS) for distributing the redundant storage of large data sets across a cluster. Use of the Hadoop framework has been demonstrated for several sequencing-related bioinformatic tasks including short read mapping \autocite{Schatz:2009p278}, SNP calling \autocite{Langmead:2009p1225}, and RNA-seq differential expression analysis \autocite{Langmead:2010p1268}.

The need to separate logic into mappers and reducers makes it difficult to implement traditional RP-based SV detection approaches in MapReduce, particularly given the global clustering of paired end mappings at the heart of many RP approaches. MapReduce algorithms, by contrast, excel at conducting many independent calculations in parallel. In sequencing applications, for example, MapReduce based SNP-callers Crossbow \autocite{Langmead:2009p1225} and GATK \autocite{McKenna:2010p1051} perform independent calculations for each location or partition in the genome. SV approaches that are similarly based on local computations have been described: the RP-based SV callers MoDIL \autocite{Lee:2009da} and forestSV \autocite{Michaelson:2012fj} attempt to solve the SV detection problem by computing scores or features along the genome and then producing SV predictions from those features in a post-processing step.

In this paper, we describe a framework for solving SV detection problems in Hadoop and MapReduce based on the computation of local features along the genome from paired end mappings. We then demonstrate the development in this framework of a software package, Cloudbreak, for discovering genomic deletions up to 25,000bp long, and short insertions. Cloudbreak computes local features that build on the ideas implemented in MoDIL \autocite{Lee:2009da}, which modeled the distribution of insert sizes at each genomic location as a Gaussian Mixture Model (GMM). Use of the Hadoop/MapReduce framework enables scalable implementations of this class of algorithm. We characterize our algorithm's performance on simulated and real data sets and compare it with those of recently published methods.

>>>>

Genomic structural variations such as deletions, insertions, and inversions of DNA are widely prevalent in human populations and account for the majority of the bases that differ among normal human genomes \autocite{Mills:2011p1611, Conrad:2010ja}. However, detection of structural variations with current high-throughput sequencing technology remains a difficult problem, with limited concordance between available algorithms and high false discovery rates \autocite{Mills:2011p1611}. Part of the problem stems from the fact that SV detection requires high coverage data sets and large computational resources. Thus, as the volume of massively parallel sequencing data approaches scales that are associated with big data, SV detection becomes a time consuming component of genomics pipelines on traditional cluster architectures, and presents a significant challenge for genomics groups or clinical operations that may not be able to scale up their computational resources readily in order to stay on top of their workloads. Here we present a distributed solution that is scalable and readily available on the cloud

In the big data realm, challenges of this scale have been addressed through distributed processing, i.e. taking the computation to the data via many compute nodes working on local chunks of data distributed over the cluster. In particular, Google's MapReduce \autocite{Dean:2008p277} framework was designed to manage the storage and efficient processing of very large scale data sets across clusters of commodity servers. Hadoop is an open source project of the Apache Foundation which provides an implementation of the MapReduce programming framework as well as a distributed file system (HDFS) for distributing the redundant storage of large data sets across a cluster. Use of the Hadoop framework has been demonstrated for several sequencing-related bioinformatic tasks including short read mapping \autocite{Schatz:2009p278}, SNP calling \autocite{Langmead:2009p1225}, and RNA-seq differential expression analysis \autocite{Langmead:2010p1268}.

@@

Our goal is to leverage the strengths of the MapReduce computational framework in order to provide fast, accurate and readily scalable SV-detection pipelines. The main challenge in this endeavor is the need to separate logic into mappers and reducers, which makes it difficult to implement traditional RP-based SV detection approaches in MapReduce, particularly given the global clustering of paired end mappings at the heart of many RP approaches. MapReduce algorithms, by contrast, excel at conducting many independent calculations in parallel. In sequencing applications, for example, MapReduce based SNP-callers Crossbow \autocite{Langmead:2009p1225} and GATK \autocite{McKenna:2010p1051} perform independent calculations for each location or partition in the genome. SV approaches that are similarly based on local computations have been described: the RP-based SV callers MoDIL \autocite{Lee:2009da} and forestSV \autocite{Michaelson:2012fj} attempt to solve the SV detection problem by computing scores or features along the genome and then producing SV predictions from those features in a post-processing step.

In this paper, we describe a framework for solving SV detection problems in Hadoop and MapReduce based on the computation of local features along the genome from paired end mappings. We then demonstrate the development in this framework of a software package, Cloudbreak, for discovering genomic deletions up to 25,000bp long, and short insertions. Cloudbreak computes local features that build on the ideas implemented in MoDIL \autocite{Lee:2009da}, which modeled the distribution of insert sizes at each genomic location as a Gaussian Mixture Model (GMM). Use of the Hadoop/MapReduce framework enables scalable implementations of this class of algorithm. We characterize our algorithm's performance on simulated and real data sets and compare it with those of recently published methods. FInally, our Cloudbreak distribution provides an automatic method to set up and configure MapReduce (Hadoop) clusters on a variety of cloud services, making distributed Sv-detection accessible to groups that may not have the resources to maintain a MapReduce capable distributed compute server.
